---
title: "DATA607 Week 9 Assignment"
subtitle: 'Web APIs'
author: "Olivia Azevedo"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(httr2)
library(keyring)
library(jsonlite)
library(dplyr)
library(kableExtra)
library(purrr)

knit_table <- function(df, caption='', position = 'left', color = '#2d3c52') {
  kbl(df, format = "html", escape = FALSE, caption = caption) |>
    kable_styling(
      full_width = F,
      position = position,
      bootstrap_options = c("striped")) |>
    row_spec(0, bold = T,
             color = "white",
             background = color,
             extra_css = paste("border: 2px solid", color, ";")) |>
    row_spec(dim(df)[1], extra_css = paste("border-bottom: 2px solid", color, ";")) |>
    column_spec(1, extra_css = paste("border-left: 2px solid", color, ";")) |>
    column_spec(dim(df)[2], extra_css = paste("border-right: 2px solid", color, ";"))
}

color_str <- function(x, color='#22a298') {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color,
      x)
  } else x
}
```


## Overview

This assignment creates an interface to pull data from the [Developer New York Times API web site](<https://developer.nytimes.com/apis>) via the `httr2` package. Specifically, the "Most Popular" API was selected to request and output the most popular articles on NYTimes.com based on emails, shares, or views. The Most Popular API JSON data response is transformed into an R DataFrame and is displayed along with supporting info on the API request.


<hr style="border:1px solid #2d3c52">


## API Request Function
Function to request article data from Most Popular NYT API from the specified previous number of days and API scope. The returned JSON response data is transformed into an R dataframe and subset by object and columns. A table of the final dataframe as well as a table with general API request and response information is displayed. The function checks to ensure response specific parameters meet the allowed parameter requirements and returns text explaining why if not.

**`r color_str('Parameters')`**:

* **api_key (character)**: user specific NYT Most Popular API Key string
* **period (int)**: number of previous days in which most popular NYT article data is returned for
  + values: 1, 7, or 30
* **scope (character)**: api scope of for getting the most popular NYT articles
  + values: 
      - viewed: most viewed articles within the last (x) day(s)
      - shared: most shared articles on Facebook within the last (x) day(s)
      - emailed: most emailed within the last (x) day(s)

**`r color_str('Returns')`**:  

  * **if correct parameters input**: dataframe of JSON API response NYT article data
    for the parameters input
  * **if wrong parameters input**: text explaining parameter error


```{r}
nyt_pop_api_resp_df = function(api_key, period, scope) {
  # define text
  df_txt = paste(
    'API Response Data - Most<b>', scope, '</b>articles for the last<b>', period, '<b>day(s)')
  info_txt = paste(
    'API Response Info - Most<b>', scope, '</b>articles for the last<b>', period, '<b>day(s)')
  
  # period and scope input error handling
  if (period %in% c(1, 7, 30) & scope %in% c('viewed', 'emailed', 'shared')) {
    request_url = paste0('https://api.nytimes.com/svc/mostpopular/v2/',scope,'/')
    
    # api request
    req = request(request_url) |>
      req_url_path_append(paste0(period,'.json')) |>
      req_url_query('api-key' = api_key) |>
      req_perform() 
    
    # apt response 
    resp = req |>
      resp_body_json(flatten = T, simplifyDataFrame=T)
    
    # create api info table
     info = tibble(
        Most_Popular_API_Status = resp$status,
        Scope = scope,
        Days_Period = period,
        Article_Count = resp$num_results,
        Source = substr(resp$copyright, 15, 45)
        )
     
    # response df data handling
    df = resp$results |>
      select(published_date, updated, title, section, subsection, adx_keywords, byline, type, url)
    
    # display results
    print(knit_table(info, caption = info_txt, color = '#557874'))
    print(
      knit_table(df, caption = df_txt) |>
      scroll_box(width = "100%", height = '500px', box_css = "border: 1px solid #FFFFFF;")
    )
    return(df)
  }
  # error return
  return(error = "INPUT ERROR: input period value is not 1, 7, or 30 or input scope value is not 'viewed', 'emailed', or 'shared'")
}
```


<hr style="border:1px solid #2d3c52">


## API Request
Calls API request function (above) for every scope and period combination and displays returned output.\
**Note: **  Execution is suspended for 12 seconds between each API request since NYT only allows for 5 requests per minute.


```{r results='asis'}
api_key = key_get('NYT_Popular_API_Key')

for (s in c('viewed', 'shared', 'emailed'
            )) {
  for (p in c(1, 7, 30
              )) {
    print(paste('<br><br><b>Scope:</b>', s, '     <b>Period:</b>', p))
    resp_df = nyt_pop_api_resp_df(api_key, p, s)
    Sys.sleep(12)
  }
}
```
