---
title: "DATA607 Project 1"
subtitle: "Data Analysis"
author: "Olivia Azevedo"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(kableExtra)

knit_table <- function(df, position = 'left') {
  kbl(df, format = "html", escape = FALSE) |>
    kable_styling(
      full_width = T,
      position = position,
      bootstrap_options = c("striped")) |>
    row_spec(0, bold = T,
             color = "white",
             background = "#2d3c52",
             extra_css = "border: 2px solid #2d3c52;") |>
  scroll_box(width = "800px", height = '250px',
             extra_css = "border-bottom: 2px solid #2d3c52; border-left: 2px solid #2d3c52; border-right: 2px solid #2d3c52;")
}
```

## Overview
The goal of this project is to transform, clean, and tidy a delimited text file with chess tournament results. The resulting data is output into a csv file containing the following columns as needed: Player Name, Player State, Total Number of Points, Player Pre-Rating, and Average Pre Chess Rating of Opponents.

<hr style="border:1px solid #2d3c52">

## Data Analysis

### Import Data
**Read in txt file as data frame**

* Define field separator character
* Define column names
```{r}
data = read.delim("https://raw.githubusercontent.com/okazevedo90/DATA607_GitHub/refs/heads/main/tournamentinfo.txt",
             header = FALSE, sep = "|", skip = 4,
             col.names = c('Num', 'Player', 'Pts', 'Round_1', 'Round_2', 'Round_3', 'Round_4', 'Round_5', 'Round_6', 'Round_7', 'NA'))
```
```{r, echo=FALSE}
knit_table(data)
```

\

### Clean Data

* Filter out not needed rows and columns
* Remove leading and trailing white space on all character type columns
```{r}
df = data |>
  filter(!str_detect(Num, "-")) |>
  mutate_if(is.character, str_trim) |>
  select(-last_col())
```
```{r, echo=FALSE}
knit_table(df)
```

\

### Tidy & Transform Data
**Make each observation a single row rather than two**

* Create subset data frame containing first row of data per player 
* Create subset data frame containing second row of data per player 
    + Extract player pre-rating value from string using regex pattern match
    + Subset to only needed columns and rename
* Concatenate subset data frames by column
```{r}
df_num = df |>
  filter(str_detect(Num, '\\d'))

df_state = df |>
  filter(!str_detect(Num, '\\d')) |>
  mutate(Pre_Rating = str_match(Player, ":\\s*(\\d*)\\S*\\s*->")[,2]) |>
  select(Num, Pre_Rating) |>
  setNames(c('State','Pre_Rating'))

df = cbind(df_num, df_state)
```
```{r, echo=FALSE}
knit_table(df)
```

\

**Make each variable a column**

* Pivot data to transform the seven chess round result columns into two columns: name and value
```{r}
df = df |>
  pivot_longer(
    cols = starts_with("Round"),
    names_to = "Round",
    values_to = 'Result',
    names_prefix = 'Round_'
  )
```
```{r, echo=FALSE}
knit_table(df)
```

\

**Prep data to allow for calculation of Average Pre Chess Rating of Opponents**

* Create subset data frame containing the player ID and pre-rating value for each player
* Extract the opponent ID value from string using regex pattern match
* Join subset data frame on player ID and opponent ID to create an opponent pre-rating value column
```{r}
rate_df = df |> 
  distinct(Num, Pre_Rating) |>
  mutate(Pre_Rating = as.integer(Pre_Rating)) |>
  setNames(c('Opponent','Opp_Pre_Rating'))

df = df |>
  mutate(Opponent = str_match(Result, "\\s*(\\d+)")[,2])

df = left_join(df, rate_df, by = join_by(Opponent))
```
```{r, echo=FALSE}
knit_table(df)
```

\

**Final Data Handling**

* Create Average Pre Chess Rating of Opponents per player column
* Subset to only needed columns and remove duplicates
* Adjust column data types
```{r}
df = df |>
  group_by(Num) |>
  mutate(Avg_Opp_Rate = round(mean(Opp_Pre_Rating, na.rm = TRUE))) |>
  ungroup() |>
  distinct(Player, State, Pts, Pre_Rating, Avg_Opp_Rate) |>
  mutate_at(c('Pts', 'Pre_Rating'), as.numeric)
```
```{r, echo=FALSE}
knit_table(df)
```

\

### Output Data
Write out data frame as csv file
```{r}
write.csv(df, "tournamentinfo_cleaned.csv", row.names = F)
```

