---
title: "DATA607_Week 2 Assignment"
subtitle: "R and SQL"
author: "Olivia Azevedo"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: true
---

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(RMySQL)
library(keyring)
library(tidyverse)
library(kableExtra)

knit_table <- function(table_name,  caption, position = 'left') {
  if (is.data.frame(table_name)) {
    x = table_name
  }
  else {
    query = paste('SELECT * FROM', table_name)
    x = dbGetQuery(con, query)
  }
  kbl(x, caption=caption, format = "html", table.attr = "style='width:30%;'") |>
    kable_styling(
      full_width = F,
      position = position,
      bootstrap_options = c("striped")) |>
    row_spec(0, bold = T,
             color = "white",
             background = "#2d3c52",
             extra_css = "border: 2px solid #2d3c52;") |>
    row_spec(dim(x)[1], extra_css = "border-bottom: 2px solid #2d3c52;") |>
    column_spec(1, extra_css = "border-left: 2px solid #2d3c52;") |>
    column_spec(dim(x)[2], extra_css = "border-right: 2px solid #2d3c52;")
}

```

## Overview
Movie rating data for six recent films was collected from five family members (raters). Each family member was asked to rate each of movies they have seen on a scale of one to five. If a rater had not seen the film in question, a NULL value was put in place of a rating.


---


## Part 1 & 2: Build Table and Store data in SQL database
The following SQL queries were used to create normalized movie rating data tables within the MySQL Workbench. The data is stored in the class (cunydata607) MySQL database in the 'olivia.azevedo90' schema.
```
-- rater table
CREATE TABLE rater (
    person_id VARCHAR(5) UNIQUE,
    person_name VARCHAR(100) UNIQUE);
INSERT INTO rater (person_id, person_name) VALUES
  ('1', 'Neil'),
  ('2', 'Myles'),
  ('3', 'Anna'),
  ('4', 'Brady'),
  ('5', 'Owen');

-- movies table
CREATE TABLE movies (
    movie_id VARCHAR(5) UNIQUE,
    movie_title VARCHAR(100) UNIQUE
);
INSERT INTO movies (movie_id, movie_title) VALUES 
	('M1', 'Barbie'),
	('M2', 'Conclave'),
	('M3', 'Oppenheimer'),
	('M4', 'Saturday Night'),
	('M5', 'Smile 2'),
	('M6', 'Wicked');

-- rates table
CREATE TABLE rates (
    person_id VARCHAR(5),
    movie_id VARCHAR(5),
    rating INT
);
INSERT INTO rates (person_id, movie_id, rating) VALUES
	-- Person 1
	('1', 'M3', 5),
	('1', 'M1', 1),
	('1', 'M5', NULL),
	('1', 'M6', 5),
	('1', 'M2', 4),
	('1', 'M4', NULL),
	-- Person 2
	('2', 'M3', NULL),
	('2', 'M1', 5),
	('2', 'M5', 5),
	('2', 'M6', NULL),
	('2', 'M2', NULL),
	('2', 'M4', 4),
	-- Person 3
	('3', 'M3', NULL),
	('3', 'M1', 3),
	('3', 'M5', 1),
	('3', 'M6', 5),
	('3', 'M2', NULL),
	('3', 'M4', 5),
	-- Person 4
	('4', 'M3', 5),
	('4', 'M1', 2),
	('4', 'M5', 3),
	('4', 'M6', NULL),
	('4', 'M2', NULL),
	('4', 'M4', NULL),
	-- Person 5
	('5', 'M3', 4),
	('5', 'M1', 3),
	('5', 'M5', 1),
	('5', 'M6', 3),
	('5', 'M2', 5),
	('5', 'M4', 4);
```

---


## Part 3: Transfer data from SQL database to R dataframe 

### Database Connection
Define connection string to MySQL database \
**Note:** To avoid hard coding database credentials, the database password was defined using the `keyring` package and is stored in the keyring store of the operating system. The stored credential can be accessed via the `key_get()` function as seen in the connection string below.
```{r}
con = dbConnect(MySQL(),
          user='olivia.azevedo90',
          password=key_get("MySQL_PASSWORD"),
          host='cuny607sql.mysql.database.azure.com',
          port=3306,
          dbname='olivia.azevedo90')
```

### Import Data
Load normalized data from the SQL database into an R data frame
```{r}
query = '
  SELECT 
  	rates.person_id,
      person_name,
      rates.movie_id,
      movie_title,
      rating
  FROM rates
  LEFT JOIN rater
  	on rates.person_id = rater.person_id 
  LEFT JOIN movies
  	on rates.movie_id = movies.movie_id'
df = dbGetQuery(con, query)
head(df)
```

---


## Part 4: Missing data strategy 

### Assess Missing Data
Identify and assess extent of missing values
```{r, echo=FALSE}
cat('Total Missing Values: ', sum(is.na(df)), '\n')

df_info = tibble(
  column = colnames(df),
  na_count = colSums(is.na(df)),
  dtype = sapply(df,class))
knit_table(df_info,  'Column NA Counts and Data Type')
```

### Missing Data Approach
My approach to addressing the movie rating missing values involves filling the NA values with zero. I choose this approach due to the missing values being explicit and representing that the person providing the rating did not see the movie.\

Removing the missing values would **not** be the most optimal approach due to:

1. A significant amount of data (30% or 10/30 rows) would be removed from the data
    + There are no rows in which all the data is missing
2. The missing values do **not** represent uncollected data or data quality issues
    + The missing values provide additional insight by indicating the rater has not seen the movie\
    
In addition, filling the values with zero converts the rating column data type from integer to numeric. Due to wanting to maintain a numerical data type, the missing values were not filled with a string.\
Lastly, the missing values were not replaced with a number other than zero (manually or via back/forward filling) since this would cause the data to be inaccurate and not reflect the true movie ratings provided by the raters.

```{r}
df_nona = df |>
  mutate(rating = coalesce(rating, 0))
```

```{r, echo=FALSE}
df_nona_info = tibble(
  column = colnames(df_nona),
  na_count = colSums(is.na(df_nona)),
  dtype = sapply(df_nona,class))
knit_table(df_nona_info,  'Column NA Counts and Data Type')
```

---


## Bonus Challenge Questions

<span style='color:#22a298'>
Are you able to use a password without having to share the password with people who are viewing your code?
</span>\
Yes - please refer to the note in the "Database Connection" section

<span style='color:#22a298'>
Can you create a normalized set of tables that corresponds to the relationship between your movie viewing friends and the movies being rated?
</span>\
I created a normalized set of tables by splitting my original table into three tables: raters, movies, and rates.

* raters
  + Columns: person_id, person_name
* movies
  + Columns: movie_id, movie_title
  + Updates: created movie id column to make the movie title dependent on a primary key
* rates
  + Columns: person_id, movie_id, rating
  + Updates: assigned the ratings to a person id and movie id to remove transitive partial dependency


### Normalized Tables

```{r, echo=FALSE}
knit_table('rater', 'Table: rater', 'float_left')
knit_table('movies',  'Table: movies', 'float_left')
knit_table('rates',  'Table: rates', 'right')
```

### Non-Normalized Table

```{r, echo=FALSE}
knit_table('movie_rates', 'Table: movie_rates', 'left')
```

---


## Exploratory Data Analysis

### Movie Rating Distribution Plots

Overall Rating Distribution
```{r}
df_nona |>
  group_by(rating) |>
  tally() |>
  ggplot(aes(x = rating, y = n)) +
    geom_col() +
    labs(
      title = "Movie Rating Distribution",
      x = 'Rating',
      y = 'Count') + 
    theme_classic()
```

Rating Distribution by Rater
```{r}
df_nona |>
  group_by(rating,  person_name) |>
  tally() |>
  ggplot(aes(x = rating, y = n, fill=person_name)) +
    geom_col(position = position_dodge2(width = 0.9, preserve = "single")) +
    labs(
      title = "Movie Rating Distribution by Rater",
      x = 'Rating',
      y = 'Count') + 
    theme_classic()
```


### Movie Rating Plots

All Movie Ratings by Rater
```{r}
df_nona |>
  filter(rating >=1) |>
  ggplot(
    aes(x = movie_title, y = rating, fill=person_name)) +
    geom_col(position = position_dodge2(width = 0.9, preserve = "single")) +
    labs(
      title = "Movie Rating Distribution by Rater",
      x = 'Movie',
      y = 'Rating') + 
    theme_classic()
```


Average Movie Rating
```{r}
df_nona |>
  group_by(movie_title) |>
  summarise(avg_ratings = mean(rating)) |>
  ggplot(
    aes(x = reorder(movie_title, -avg_ratings), y = avg_ratings)) +
    geom_col(fill='#2d3c52') +
    labs(
      title = "Average Movie Rating",
      x = 'Movie',
      y = 'Average Rating') + 
    theme_classic()
```

Average Movie Rating by Rater
```{r}
df_nona |>
  group_by(person_name) |>
  summarise(avg_ratings = mean(rating)) |>
  ggplot(
    aes(x = reorder(person_name, -avg_ratings), y = avg_ratings)) +
    geom_col(fill='#557874') +
    labs(
      title = "Rater Average Movie Rating",
      x = 'Rater',
      y = 'Average Rating') + 
    theme_classic()
```



### Movies Watched Plots

Movies Watched Distribution
```{r}
df_nona |>
  filter(rating >=1) |>
  group_by(movie_title) |>
  tally() |>
  ggplot(
    aes(x = reorder(movie_title, -n), y = n)) +
    geom_col(fill='#2d3c52') +
    labs(
      title = "Movies Watched Distribution",
      x = 'Movie',
      y = 'Watched Count') + 
    theme_classic()
```

Movies Watched by Rater Distribution
```{r}
df_rater_n = df_nona |>
  filter(rating >=1) |>
  group_by(person_name) |>
  tally()

  ggplot(df_rater_n,
    aes(x = reorder(person_name, -n), y = n)) +
    geom_col(fill='#557874') +
    labs(
      title = "Movies Watched by Rater Distribution",
      x = 'Movie',
      y = 'Watched Count') + 
    theme_classic()
```

### Movie Rating & Movies Watched Plot

Average Rating per the Number of Movies Watched by the Rater
```{r}
merge(df_nona, df_rater_n) |>
  group_by(n) |>
  summarise(avg_ratings = mean(rating)) |>
  ggplot(
    aes(x = reorder(n, -avg_ratings), y = avg_ratings)) +
  geom_col(fill='#2d3c52') +
  labs(
      title = "Average Rating per Number of Movies Watched",
      x = "Number of Movie's Watched",
      y = 'Average Rating') + 
    theme_classic()
```

---

## Conclusions
Due to this data set being so small and having many missing rating values, there is not enough evidence to draw any meaningful conclusions. Some potential insights include:

* The number of movie's watched being positively correlated to movie ratings
* The movie Barbie being the most highly rated movie of the six movies included in the survey
  + **Note:** This is most likely influenced by the movie Barbie being the only movie that all raters watched and provided a rating for

To improve this analysis, I would recommend collecting more movie ratings via surveying a much larger population. A much larger data set will provide a better representation of overall movie ratings, as well as provide opportunity to apply a different approach to addressing missing rating values. 

