---
title: "DATA607_Week 3 Assignment"
subtitle: 'Normalization & Character Manipulation'
author: "Olivia Azevedo"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: true
---

<hr style="border:1px solid #2d3c52">

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(kableExtra)
library(randomNames)

color_str <- function(x, color='#2d3c52') {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color,
      x)
  } else x
}

knit_table <- function(df,  caption, position = 'left') {
  kbl(df, caption=caption, format = "html", table.attr = "style='width:30%;'") |>
    kable_styling(
      full_width = F,
      position = position,
      bootstrap_options = c("striped", "condensed")) |>
    row_spec(0, bold = T,
             color = "white",
             background = "#2d3c52",
             extra_css = "border: 2px solid #2d3c52;") |>
    row_spec(dim(df)[1], extra_css = "border-bottom: 2px solid #2d3c52;") |>
    column_spec(1, extra_css = "border-left: 2px solid #2d3c52;") |>
    column_spec(dim(df)[2], extra_css = "border-right: 2px solid #2d3c52;")
}
```

## Overview
The following assignment demonstrates data normalization and character manipulation. Data normalization organizes data in a consistent way and to reduce redundancy and ensure data integrity. Character manipulation is an essential skill needed to clean and transform character data.

<hr style="border:1px solid #2d3c52">

## Normalization

### Question 1

College course data was synthetically generated to create four dataframes that demonstrate normalization. The dataframes include student, course, department, and teacher related info in third normal form.
A seed was set to ensure reproducible generated data results .

```{r}
set.seed(789)
student_df = data.frame(
  STUDENT_ID = sprintf("S%02d", 1:30),
  COURSE_ID = sample(c("D101", "D102", "D103", "N101", "N102", "N103"), 30, replace = TRUE),
  STUDENT_FIRST_NAME = randomNames(30, which.names="first"),
  STUDENT_LAST_NAME = randomNames(30, which.names="last")
)

course_df = data.frame(
  COURSE_ID = c("D101", "D102", "D103", "N101", "N102", "N103"),
  DEPARTMENT_ID = c("DEPT1", "DEPT1", "DEPT1", "DEPT2", "DEPT2", "DEPT2"),
  TEACHER_ID = c("T1", "T1", "T2", "T3", "T4", "T5"),
  COURSE_NAME = c("Data Analytics", "Stats", "Machine Learning","Chemistry", "Biology", "Pharmacology")
)

teacher_df = data.frame(
  TEACHER_ID =c ("T1", "T2", "T3", "T4", "T5"),
  TEACHER_FIRST_NAME = randomNames(5, which.names="first"),
  TEACHER_LAST_NAME = randomNames(5, which.names="last"),
  TEACHER_CONTACT = sample(1234567:87654321, 5)
)

department_df = data.frame(
  DEPARTMENT_ID = c("DEPT1", "DEPT2"),
  DEPARTMENT_NAME = c("Data Analytics", "Nursing")
)

join_df = left_join(student_df, course_df, by = join_by(COURSE_ID))
join_df = left_join(join_df, teacher_df, by = join_by(TEACHER_ID))
join_df = left_join(join_df, department_df, by = join_by(DEPARTMENT_ID))
join_df = join_df |>
  select(STUDENT_ID, STUDENT_FIRST_NAME, STUDENT_LAST_NAME,
         COURSE_ID, COURSE_NAME, TEACHER_ID, TEACHER_FIRST_NAME,
         TEACHER_LAST_NAME, TEACHER_CONTACT, DEPARTMENT_ID, DEPARTMENT_NAME)
```

#### Normalized Dataframes
```{r, echo=FALSE}
knit_table(student_df, 'Student Dataframe', 'left')
knit_table(course_df, 'Course Dataframe', 'left')
knit_table(teacher_df, 'Teacher Dataframe', 'left')
knit_table(department_df, 'Department Dataframe', 'left')
```

#### Joined Normalized Dataframes
```{r, echo=FALSE}
knit_table(join_df, 'All Joined Dataframe (scroll to view all columns and rows)', 'left') |>
  scroll_box(width = "800px", height = '600px', box_css = "border: 0px solid #000000; padding: 0px; ")
```


<hr style="border:1px solid #2d3c52">

## Character Manipulation

### Question 2

#### Import Data 
Read in college majors csv file as data frame
```{r}
cm_df = read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/refs/heads/master/college-majors/majors-list.csv")
```
\
#### College Major String Pattern Match
Identifies the majors that contain either "DATA" or "STATISTICS"
```{r}
pat_df = cm_df |> 
  filter(str_detect(Major, "DATA|STATISTICS")) |>
  select(Major)
colnames(pat_df) = c('Majors that contain either \"DATA\" or \"STATISTICS\"')
```

```{r, echo=F}
knit_table(pat_df, '')
```


---


### Question 3

Descriptions for five regular expressions have been provided below. Each expression description includes a write up on what the expression will match with supporting examples.


#### `(.)\1\1`
This regex will match strings that contain 3 identical consecutive characters.

* **`r color_str('Breakdown: ', '#22a298')`**
  + `(.)` matches any single character
  + `\1` is a back reference to the first captured group
* **`r color_str('Examples: ', '#22a298')`** '111', 'aaa'\
\

#### `"(.)(.)\\2\\1"`
This is a string that represents a regex which matches strings with any 2 characters followed by the second character and first character repeated in that order.

* **`r color_str('Breakdown: ', '#22a298')`**
  + The first `(.)` matches any single character and represents the first captured group
  + The second `(.)` again matches any single character and represents the second captured group
  + These captured groups are then back referenced with the `\\2` referring to the second captured group and `\\1` referring to the first captured group.
* **`r color_str('Examples: ', '#22a298')`** '1221', 'abba', '3333'\
\

#### `(..)\1`
This regex will match strings with a repeated pair of characters.

* **`r color_str('Breakdown: ' ,'#22a298')`**
  + `(..)`  matches any 2 characters and represents the first captured group
  + `\1` is a back reference to the first captured group.
* **`r color_str('Examples: ' ,'#22a298')`** '1212', 'abab', 'cccc'\
\

#### `"(.).\\1.\\1"`
This is a string that represents a regex which matches strings with any character repeated in 3 places if a single character is in between the repeated characters.

* **`r color_str('Breakdown: ' ,'#22a298')`**
  + `(.)` matches any single character and represents the first captured group
  + `.`   matches any single character
  + `\\1` is a back reference referring to the first captured group
* **`r color_str('Examples: ', '#22a298')`** '12131', 'abaca', 'dddd'\
\


#### `"(.)(.)(.).*\\3\\2\\1"`
This is a string that represents a regex which matches strings with any 3 characters followed by an optional character that can be repeated 0 or more times. The pattern ends with same first 3 characters being repeated in reverse order. 

* **`r color_str('Breakdown: ' ,'#22a298')`**
  + `(.)` matches any single character and creates a captured group. This is repeated 3 times; thus, three captured groups are defined
  + `.*`  optionally matches any character with a repeat (i.e. it matches any number of times, including 0).
  + `\\3\\2\\1` are back references referring to the third, second, and first captured group characters respectively 
* **`r color_str('Examples: ' ,'#22a298')`** '123321', 'abcdddddddcba'\

---



### Question 4

Regex to match words that start and end with the same character\
**`r color_str('Regex: ' ,'#22a298')`** `^(.).*\\1$`

Test on `words` character vector
```{r}
str_subset(words, "^(.).*\\1$")
```
Regex to match words that contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.)\
**`r color_str('Regex: ' ,'#22a298')`** `([a-zA-Z]{2}).*\\1`

Test on `words` character vector
```{r}
str_subset(words, "([a-zA-Z]{2}).*\\1")
```

Regex to match words that contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.)\
**`r color_str('Regex: ' ,'#22a298')`** `([a-zA-Z])(.*\\1){2,}`

Test on `words` character vector
```{r}
str_subset(words, "([a-zA-Z])(.*\\1){2,}")
```
